A primer to numerical simulations: The perihelion motion of Mercury

### 2018-01-19 tex/MercuryNumerical.tex
added summary (@CH)

\begin{abstract}
Numerical simulations are playing an increasingly important role in modern science.
In this work it is suggested to use a numerical study of the famous perihelion motion of the planet Mercury (one of the prime observables supporting Einsteins General Relativity) as a test case to teach numerical simulations to high school students.
The paper includes details about the development of the code as well as a discussion of the visualization of the results.
In addition a method is discussed how to estimate the size of the effect a priori. Which allows to double check the results found numerically.
\end{abstract}

@@

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise. This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to learn about
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \vpython{}.
\end{itemize}
We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
Their detailed study needs a deep knowledge of Differential Geometry. For the study outlined here, however, very little math is necessary, such that high school students familiar to derivatives benefit from it.
This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.

The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their
discretization is discussed based on Newton's law of gravitation and possible extensions thereof.
Afterwards the visualization of the resulting trajectories using \vpython{} is introduced and applied to the problem at hand. In addition,
 tools are developed to extract the relevant quantity from the result of the simulation.
Finally, the principle of dimensional analysis is presented as a tool not only to estimate some effect but also to cross check whether the results of the simulation are sensible.

@@

A physical system is said to be understood if the assumed forces acting on it lead to the observed trajectories.
In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space), the location is parametrized by a single three vector $\vec r(t)$.
Additionally, we need to be able to calculate the objects velocity $\vec v(t)$ and acceleration $a(t)$ in order to describe and control its dynamics; this will become clear below. The velocity describes a change of location over time which, for some infinitesimally small $\Delta t$, can be written as
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + \ldots \ . \label{eq:vdef}
\end{equation}
Similarly, the acceleration describes a change in velocity, i.e.
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + \ldots\ . \label{eq:adef}
\end{eqnarray}
The dots in these expressions indicate additional terms that may be expressed in terms of higher powers in $\Delta t$. While they are needed in general, for sufficiently small $\Delta t$ they can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{d t}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced a common short hand notation for time derivatives in the last expression.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced the second derivative in the second line.

It was Newton who observed that if a body is at rest, it will remain at rest, and if it is in motion it will remain in motion at a constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newton's first law.
Formulated differently: a force $\vec F$ expresses itself by changing the motion of some object.
This is quantified in Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation}
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as fuel is burned.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation}
Note that in general the force could depend also on the time or the velocity.
Here, we restrict ourselves to the case relevant for our example, where the force depends on the location only.
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g., solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge of mathematics and sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use the force to calculate $\vec a(t) = \vec F(\vec r(t)) / m$, see Eq.~(\ref{eq:newton2}).
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$.
\item Use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
	In general, a differential equation of $n^{\rm th}$ degree (where the highest derivative is order $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some starting time, but one may as well pick two locations at different times.%
}.

This procedure can only work if $\Delta t$ is sufficiently small.
What this means depends on the process one studies.
One way to estimate whether $\Delta t$ is small enough, is to verify whether the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\
\label{eq:check}
\end{equation}
holds. This follows from Eq.~(\ref{eq:vdef}) where the first term that we neglected reads $(1/2)a(t){(\Delta t)}^2$.
Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newton's law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=1.99\times 10^{30}$ kg is the mass of the Sun.
In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
Although this is not exact, it is a pretty good approximation because $m/M_\odot\sim 10^{-8}$.
For later convenience we introduce the Schwarzschild radius of the Sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 2.95 \ \mbox{km} \ , \label{rsdef}
\end{equation}
where $c=3.00\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the Sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
The Schwarzschild radius is also an important quantity to characterize black holes; this is however not relevant for the discussion at hand.
With this, Newton's second law reads
\begin{equation}
\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, . \label{eq:newton}
\end{equation}
An attractive force that vanishes at large distances leads in general, depending on the initial conditions, either to bounded or to open orbits.
In the latter case the planet simply disappears from the Sun, since a given force can capture only bodies with small enough momenta.
The students may study those scenarios within their simulations by varying the start velocity while keeping the start location fixed.

@@

The observed perihelion motion of Mercury is nowadays determined as
\[(574.10\pm 0.65)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}}\]
where the symbol $''$ denotes "arc seconds": 1$''={(1/3600)}^o$. The bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of
\begin{equation}
\delta \Theta_M = (43)'' \ \mbox{in 100 earth years} \label{delT}
\end{equation}
remained unexplained, until Einstein quantified the predictions of General Relativity to this particular observable~\cite{Einstein}.

To allow for a movement of the perihelion we need to modify the force that led to Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter.
This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from an $1/r$ potential expressed relative to the characteristic length $r_S$.
However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
\begin{equation}
r_a^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $r_a^2$ carries dimensions of length squared.
We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta (r_a^2/r^2)$.
Combining everything, we use the following ansatz for the modified equation of motion:
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand, $r_a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
 $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59.0$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
We thus find that $(r_a^2/ r_{MS}^2) \sim 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})\sim6\cdot 10^{-8}$.
Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect of similar size. Furthermore, any term of higher order in either $a$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data --- this will be done below.
However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
General relativity gives~\cite{Einstein}
\begin{equation}
	\alpha = 0 \, , \quad \beta = 3 \, . \label{eq:art-prediction}
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).

@@

If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in figure~\ref{fig:MercuryOrbit-a0-small}.
At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also figure~\ref{fig:MercuryOrbit-a0-small-dt-large}].
To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing  $\alpha\neq0$
[figure~\ref{fig:MercuryOrbit-a6-small}].
This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories. In particular
they should discover that values of $\alpha$ of the order of $10^6$ are necessary to get a visible effect.
For a nicely visible perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [figure~\ref{fig:MercuryOrbit-a6-big}].
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/num-err.png}
		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit}
		Different Mercury orbits for $\beta=0$ and
		\begin{itemize}
		\item[(a)] $\alpha=0$ and $\Delta t = \Delta t_0  / 20$;
		\item[(b)] $\alpha=0$ and $\Delta t = \Delta t_0  \times 2$, where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
		\item[(d)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
		\end{itemize}
		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) /a_M(0)$ and the images are screenshots of the simulation (with modified colors).
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "Mercury year" changes.
	}
\end{figure}

@@

%\textit{What exactly do we want the students / advise the teachers to do? I think a nice approach would be to let the students implement the code to extract the perihelion motion and make a list of some values of alpha and the corresponding angles. Then discuss, that $\alpha=1$ is the natural size for $\alpha$ and that to extract the value of angle they need to exploit the linear dependence. Finally they can think on whether the value they get is reasonable and the teacher can explain the rest of section? What do you think?}\\
Both $\alpha$ and $\beta$ have similar effects on the perihelion motion.
We recommend setting one of them to zero and varying the other when analyzing the perihelion motion in dependence on the parameters.
In order to calculate the size of the perihelion motion for a modified gravitational force, the students need to
\begin{itemize}
\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and $\beta$. The  simulation time needs to cover several revolutions.
\item calculate the angle between the perihelions for each pair of successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles.
\item repeat the steps given above for different values of $\alpha$ (or $\beta$). The students should find a linear dependence of $\delta \Theta$ on $\alpha$ or $\beta$.
\item interpolate $\delta \Theta (\alpha, \beta)$ to small values of $\alpha$ (or $\beta$) --- including  $\delta \Theta(0,0)=0$.
\end{itemize}

@@

\todo[inline]{Summary}

>>>>

\begin{abstract}
Numerical simulations are playing an increasingly important role in modern science.
In this work it is suggested to use a numerical study of the famous perihelion motion of the planet Mercury (one of the prime observables supporting Einsteins General Relativity) as a test case to teach numerical simulations to high school students.
The paper includes details about the development of the code as well as a discussion of the visualization of the results.
In addition a method is discussed that allows one to estimate the size of the effect as well as the uncertainty of 
the approach a priori. At the same time this enables the students to double check the results found numerically.
\end{abstract}

@@

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise. This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
This paper introduces a project that allows one to demonstrate the power of numerical simulations.
On the example of the perihelion motion of the planet Mercury the students are supposed to learn about
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic tests and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using \vpython{}.
\end{itemize}
We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
Their detailed study needs a deep understanding of Differential Geometry. For the course presented here, however, very little math is necessary, such that high school students familiar to derivatives will benefit from it.
This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.

The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their
discretization is discussed based on Newton's law of gravitation and possible extensions thereof.
Afterwards the visualization of the resulting trajectories using \vpython{} is introduced and applied to the problem at hand. In addition,
 tools are developed to extract the relevant quantity from the result of the simulation.
Finally, the principle of dimensional analysis is presented as a tool not only to estimate some effect as well as the 
expected accuracy of a calculation, but also to cross check whether the results of the simulation are sensible.
As is pointed out in the chapters below the course can be finished at some canonical points --- in particular we think
that for most students a detailed study of the uncertainties of the numerical simulations might be too technical.
However, we put also this discussion into the paper not only for completeness but also to provide an additional 
challenge to the more advanced students.

@@

A physical system is said to be understood, if the assumed forces acting on it lead to the observed trajectories.
In other words, we need to show that we can calculate the location in space of the object of interest at any future point in time, once the initial conditions are fixed properly.
If we can neglect the finite size of this object (and in particular its orientation in space), the location is parametrized by a single three vector $\vec r(t)$.
Additionally, we need to be able to calculate the objects velocity $\vec v(t)$ and acceleration $\vec a(t)$ in order to describe and control its dynamics; this will become clear below. The velocity describes a change of location over time which, for some infinitesimally small $\Delta t$, can be written as
\begin{equation}\label{eq:x_update}
\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + \ldots \ . \label{eq:vdef}
\end{equation}
Similarly, the acceleration describes a change in velocity, i.e.
\begin{eqnarray}\label{eq:v_update}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + \ldots\ . \label{eq:adef}
\end{eqnarray}
The dots in these expressions indicate additional terms that may be expressed in terms of higher powers in $\Delta t$. While they are needed in general, for sufficiently small $\Delta t$ they can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{d t}  = \dot{\vec  r}(t) \ ,
\end{equation}
where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced a common short hand notation for time derivatives in the last expression.
Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
\end{eqnarray}
where we introduced the second derivative in the second line.

It was Newton who observed that, if a body is at rest, it will remain at rest, and if it is in motion it will remain in motion at a constant velocity in a straight line, 
unless it is acted upon by some force --- this is known as Newton's first law.
Formulated differently: a force $\vec F$ expresses itself by changing the motion of some object.
This is quantified in Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation}
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as fuel is burned.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation}
Note that in general the force could depend also on the time or the velocity.
Here, we restrict ourselves to the case relevant for our example, where the force depends on the location only.
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory by solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge of mathematics and sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use the force to calculate $\vec a(t) = \vec F(\vec r(t)) / m$, see Eq.~(\ref{eq:newton2}).
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$.
\item Use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
	In general, a differential equation of $n^{\rm th}$ degree (where the highest derivative is order $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some starting time, but one may as well pick two locations at different times.%
}.

This procedure can only work, if $\Delta t$ is sufficiently small.
What this means depends on the process one studies.
One way to estimate whether $\Delta t$ is small enough, is to verify whether the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\
\label{eq:check}
\end{equation}
holds. This follows from Eq.~(\ref{eq:vdef}) where the first term that we neglected reads $(1/2)a(t){(\Delta t)}^2$.
Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly. This issue will
be discussed below in more detail.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newton's law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=1.99\times 10^{30}$ kg is the mass of the Sun.
In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury, when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
Although this is not exact, it is a pretty good approximation because $m/M_\odot\sim 10^{-8}$.
For later convenience we introduce the Schwarzschild radius of the Sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 2.95 \ \mbox{km} \ , \label{rsdef}
\end{equation}
where $c=3.00\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the Sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
The Schwarzschild radius is also an important quantity to characterize black holes; this is however not relevant for the discussion at hand.
With this, Newton's second law reads
\begin{equation}
\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, . \label{eq:newton}
\end{equation}
An attractive force that vanishes at large distances leads in general, depending on the initial conditions, either to bounded or to open orbits.
In the latter case the planet simply disappears from the Sun, since a given force can capture only bodies with small enough momenta.
The students may study those scenarios within their simulations by varying the start velocity while keeping the start location fixed.

@@

The observed perihelion motion of Mercury is nowadays determined as
\[(574.10\pm 0.65)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}}\]
where the symbol $''$ denotes "arc seconds": 1$''={(1/3600)}^o$. The bulk of this number
can be understood by the presence of other planets within the Newtonian theory, since
their gravitational force also acts on Mercury. However, a
residual motion of
\begin{equation}
\delta \Theta_M = (42.56\pm 0.94)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}} \label{delT}
\end{equation}
remained unexplained, until Einstein quantified the predictions of General Relativity to this particular observable~\cite{Einstein}.

To allow for a movement of the perihelion we need to modify the force that led to Eq.~(\ref{eq:newton}).
This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter.
This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from an $1/r$ potential expressed relative to the characteristic length $r_S$.
However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
\begin{equation}
r_a^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
\end{equation}
where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
It is easily verified that $r_a^2$ carries dimensions of length squared.
We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta (r_a^2/r^2)$.
Combining everything, we use the following ansatz for the modified equation of motion:
\begin{equation}
\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
\label{eq:newton_art}
\end{equation}
For the system at hand, $r_a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
 $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59.0$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
$ r_{MS}=\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6$~km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
We thus find that $(r_a^2/ r_{MS}^2) \sim 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})\sim6\cdot 10^{-8}$.
Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect of similar size. Furthermore, any term of higher order in either $a$ or $r_S$
should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
The parameters $\alpha$ and $\beta$ can be extracted from a fit to data.
However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
General Relativity gives~\cite{Einstein}
\begin{equation}
	\alpha = 0 \, , \quad \beta = 3 \, . \label{eq:art-prediction}
\end{equation}
Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
values given in Eq.~(\ref{eq:art-prediction}).

@@

If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in figure~\ref{fig:MercuryOrbit-a0-small}.
At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also figure~\ref{fig:MercuryOrbit-a0-small-dt-large}].
To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing  $\alpha\neq0$
[figure~\ref{fig:MercuryOrbit-a6-small}].
This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories. In particular
they should discover that values of $\alpha$ of the order of $10^6$ are necessary to get a nicely visible effect.
For an even more enhanced perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [figure~\ref{fig:MercuryOrbit-a6-big}].
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/num-err.png}
		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
		\caption{\label{fig:MercuryOrbit-a6-big}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit}
		Different Mercury orbits for $\beta=0$ and
		\begin{itemize}
		\item[(a)] $\alpha=0$ and $\Delta t = \Delta t_0  / 20$;
		\item[(b)] $\alpha=0$ and $\Delta t = \Delta t_0  \times 2$, where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
		\item[(d)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
		\end{itemize}
		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) /a_M(0)$ and the images are screenshots of the simulation (with modified colors).
		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "Mercury year" changes.
	}
\end{figure}

This finalizes that basic course ---
for the not so ambitious students this might be a good point to stop. Already up to here the students should have
learned a lot about numerical physics, but the load of details was hopefully not too high. Moreover they have already the nice
achievement that they were able to simulate numerically a perihelion motion. The chapters to come are increasingly technical
and thus address more advanced students.

@@

%\textit{What exactly do we want the students / advise the teachers to do? I think a nice approach would be to let the students implement the code to extract the perihelion motion and make a list of some values of alpha and the corresponding angles. Then discuss, that $\alpha=1$ is the natural size for $\alpha$ and that to extract the value of angle they need to exploit the linear dependence. Finally they can think on whether the value they get is reasonable and the teacher can explain the rest of section? What do you think?}\\
Both $\alpha$ and $\beta$ have similar effects on the perihelion motion.
We recommend to set one of them to zero and to vary the other, when analyzing the perihelion motion. 
In order to calculate the size of the perihelion motion for a modified gravitational force, the students need to
\begin{itemize}
\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and $\beta$. The  simulation time needs to cover several revolutions.
\item calculate the angle between the perihelions for each pair of successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles.
\item repeat the steps given above for different values of $\alpha$ (or $\beta$). The students should find a linear dependence of $\delta \Theta$ on $\alpha$ or $\beta$.
\item interpolate $\delta \Theta (\alpha, \beta)$ to small values of $\alpha$ (or $\beta$) --- including  $\delta \Theta(0,0)=0$.
\end{itemize}

@@

\section{Summary}
